---
title: "Analyzing circadian transcriptome data with LimoRhyde"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing circadian transcriptome data with LimoRhyde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', fig.retina = 2, message = FALSE)
```

LimoRhyde is a framework for differential analysis of rhythmic transcriptome data. This vignette goes through the typical steps of an analysis: identifying rhythmic genes, identifying differentially rhythmic genes, and identifying differentially expressed genes.

The dataset is based on total RNA from livers of wild-type and Rev-erb$\alpha/\beta$ knockout mice, with gene expression measured by microarray ([GSE34018](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE34018)).

## Load packages and set parameters

```{r}
library('annotate')
library('data.table')
library('foreach')
library('ggplot2')
library('limma')
library('limorhyde')
library('org.Mm.eg.db')
library('qs')
```

Here we specify the zeitgeber period and the q-value cutoffs for rhythmic and differentially rhythmic genes.
```{r}
period = 24
qvalRhyCutoff = 0.15
qvalDrCutoff = 0.1
```

## Load the data

The expression data are in a matrix with one row per gene and one column per sample. The metadata are in a table with one row per sample.
```{r}
y = qread(system.file('extdata', 'GSE34018_expression_data.qs', package = 'limorhyde'))
y[1:5, 1:5]

metadata = qread(system.file('extdata', 'GSE34018_metadata.qs', package = 'limorhyde'))
metadata
```

We use `limorhyde` to calculate `time_cos` and `time_sin`, which are based on the first harmonic of a Fourier decomposition of the `time` column, and append them to the `metadata` data.table.
```{r}
metadata = cbind(metadata, limorhyde(metadata$time, 'time_'))
```

## Identify rhythmic genes

The next three steps use [limma](https://doi.org/doi:10.18129/B9.bioc.limma). We calculate the q-value of rhythmicity for each gene using that gene's p-value for each condition and adjusting for multiple testing.
```{r}
rhyLimma = foreach(condNow = unique(metadata$cond), .combine = rbind) %do% {
  design = model.matrix(~ time_cos + time_sin, data = metadata[cond == condNow])
  fit = lmFit(y[, metadata$cond == condNow], design)
  fit = eBayes(fit, trend = TRUE)
  rhyNow = data.table(topTable(fit, coef = 2:3, number = Inf), keep.rownames = TRUE)
  setnames(rhyNow, 'rn', 'gene_id')
  rhyNow[, cond := condNow]}

rhyLimmaSummary = rhyLimma[, .(P.Value = min(P.Value)), by = gene_id]
rhyLimmaSummary[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(rhyLimmaSummary, 'adj.P.Val')

rhyLimmaSummary[1:5, ]
```

## Identify differentially rhythmic genes

Differential rhythmicity is based on statistical interactions between `cond` and the `time` components. We pass all genes to limma (whose Empirical Bayes does best with many genes), but keep results only for rhythmic genes, and adjust for multiple testing accordingly.
```{r}
design = model.matrix(~ cond * (time_cos + time_sin), data = metadata)

fit = lmFit(y, design)
fit = eBayes(fit, trend = TRUE)
drLimma = data.table(topTable(fit, coef = 5:6, number = Inf), keep.rownames = TRUE)
setnames(drLimma, 'rn', 'gene_id')

drLimma = drLimma[gene_id %in% rhyLimmaSummary[adj.P.Val <= qvalRhyCutoff]$gene_id]
drLimma[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(drLimma, 'adj.P.Val')

drLimma[1:5, ]
```

## Identify differentially expressed genes

Differential expression is based on the coefficient for `cond` in a linear model with no interaction terms. We pass all genes to limma, but keep results only for non-differentially rhythmic genes, and adjust for multiple testing accordingly.
```{r}
design = model.matrix(~ cond + time_cos + time_sin, data = metadata)

fit = lmFit(y, design)
fit = eBayes(fit, trend = TRUE)
deLimma = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma, 'rn', 'gene_id')

deLimma = deLimma[!(gene_id %in% drLimma[adj.P.Val <= qvalDrCutoff]$gene_id)]
deLimma[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma, 'adj.P.Val')

deLimma[1:5, ]
```

## Plot the results

Numerous plots are possible. One is a volcano plot of differentially expressed genes.
```{r, fig.width = 4, fig.height = 3}
ggplot(deLimma) +
  geom_point(aes(x = logFC, y = -log10(adj.P.Val)), size = 0.2, alpha = 0.5) +
  labs(x = expression(log[2] * ' fold-change'), y = expression(-log[10] * ' ' * q[DE]))
```

Another is a plot of expression as a function of time and genotype for individual genes. Here we show, from top to bottom, the top rhythmic gene, top differentially rhythmic gene, and top differentially expressed gene by q-value.
```{r, fig.width = 5, fig.height = 4}
geneIdsNow = c(rhyLimmaSummary$gene_id[1L], drLimma$gene_id[1L], deLimma$gene_id[1L])
geneSymbolsNow = unname(getSYMBOL(geneIdsNow, 'org.Mm.eg.db'))

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, sample_id := colnames(y[geneIdsNow, ])]

df = merge(df, metadata[, .(sample_id, cond, time)], by = 'sample_id')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]

ggplot(df) +
  facet_grid(symbol ~ cond, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = cond, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (norm.)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))
```
